The \dealii Library, Version 9.2

### 2020-05-23 9.2/paper.tex
Add section for SD::BatchOptimizer
Add myself to the author list
Improve and shorten some text

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
%     Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
    Peter Munch,
%     Ross Maguire Kynch,
    Matthias Maier,
%     Jean-Paul Pelteret,
%     Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

The class \texttt{VectorizedArray<Number>} is a key ingredient for the high 
node-level performance of the matrix-free algorithms in deal.II~\cite{KronbichlerKormann2012, KronbichlerKormann2019}. It is a wrapper 
class around $n$ vector entries of type \texttt{Number} and delegates relevant 
function calls to appropriate Intrinsics instructions. Up to release 9.1, the 
vector length $n$ has been set at compile time of the library to the highest 
possible value supported by the given processor architecture.

The class \texttt{VectorizedArray} has been made more user-friendly by making 
it compatible with the STL algorithms found in the header \texttt{<algorithm>}. 
The length of the  vector can now be queried by  \texttt{VectorizedArray::size()} and its underlying number type by \texttt{VectorizedArray::value\_type}. 
Furthermore, the \texttt{VectorizedArray} class supports range-based iterations over its entries and, i.a., the following 
algorithms: \texttt{std::\allowbreak ad\-vance()}, \texttt{std::distance()}, and \texttt{std::max\_element()}.

This class has been also extended with the second optional template argument 
\texttt{VectorizedArray<Number, size>} with \texttt{size} being related to the 
vector length, i.e., the number of lanes to be used and the instruction set to be 
used. By default, the number is set to the highest value supported by the given 
hardware. A full list of supported 
vector lengths is presented in Table~\ref{tab:vectorizedarray}.

All matrix-free related classes (like \texttt{MatrixFree} and \texttt{FEEvaluation}) 
have been templated with the floating-point number type \texttt{Number} (e.g. \texttt{double} or \texttt{float}); the computations were performed implicitly 
on \texttt{VectorizedArray<Number>} structs with the highest 
instruction-set-architecture extension, with each lane responsible for a separate 
cell (vectoriziation over elements). In release 9.2, all matrix-free classes 
have been extended with a new optional template argument specifying the 
\texttt{VectorizedArrayType}. This allows users to select the vector length/ISA and, 
as a consequence, the number of cells to be processed at once directly in their applications: 
The deal.II-based 
library \texttt{hyper.deal}~\cite{munch2020hyperdeal}, which solves the 6D Vlasov--Poisson equation with high-order 
discontinuous Galerkin methods (with more than 1024 degrees of freedom per cell), works with \texttt{MatrixFree} objects of different SIMD-vector 
length in the same application and benefits---in terms of performance---by the possibility of decreasing the number of cells processed by a single SIMD instruction.

@@

A side effect of introducing the new template argument \texttt{VectorizedArrayType} 
in the \texttt{MatrixFree} classes is that any data structures 
\texttt{VectorizedArrayType} can be processed if they support required 
functionalities like \texttt{size()} or \texttt{value\_type}. In this context, we 
would like to highlight that the new \texttt{C++20} feature \texttt{std::simd} 
can be processed by the matrix-free infrastructure with minor internal 
adjustment as an open pull request shows 
(see \url{https://github.com/dealii/dealii/pull/9994}).  
Table~\ref{tab:simd} gives a comparison of the deal.II-specific SIMD classes and 
the equivalent C++20 classes. 
%We welcome the standardization of the SIMD 
%parallelization paradigm in C++ and intend to replace step by step our own 
%wrapper class, which has been continuously developed over the last decade. We 
%would like to emphasize that the work invested in this class was not in vain, 
%since many performance-relevant utility functions implemented with \texttt{VectorizedArray} in mind (e.g., \texttt{vectorized\_load\_and\_transpose} 
%and \texttt{vectorized\_transpose\_and\_store}) will be still used, since they 
%have not become part of the standard.

@@

Large-scale simulations with 304,128 cores have revealed bottlenecks in release 
9.1 during initialization due to the usage of expensive collective operations 
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak Alltoall()}, e.g., during the 
pre-computation of the index ranges of all processes, which have been stored in  an array. 
This information is needed to set up 
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class. 
In release 9.2, we have removed such arrays and have replaced 
the \texttt{MPI\_Allgather}/\allowbreak\texttt{MPI\_\allowbreak Alltoall} 
function calls by consensus algorithms~\cite{hoefler2010scalable}, which can be 
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}): now, only the locally relevant information about the index ranges is (re)computed when needed, using these algorithms. 
We provide two flavors of the consensus algorithm: the two-step 
approach \texttt{CA::PEX} and the \texttt{CA::NBX}, 
which uses only point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

Users can apply the new algorithms for their own dynamic-sparse problems by 
providing a list of target 
processes and pack/unpack routines either by implementing the interface 
\texttt{CA::\allowbreak Process} or by providing \texttt{std::function} 
objects to \texttt{CA::AnonymousProcess}.

By replacing the collective communications during setup and removing the arrays 
that contain information for each process (enabled by the application of consensus 
algorithms and other modifications---a full list of modifications leading to this 
improvement can be found online), we were able to significantly improve the setup 
time for large-scale simulations and to solve a Poisson problem with multigrid 
with 12T unknowns. 
Figure~\ref{} compares the timings of a simulation (incl. setup) with the 
previous release 9.1 and with the current release 9.2. 
{\color{red}TODO[Peter/Martin] description of the results}

>>>>

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
%     Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
    Peter Munch,
%     Ross Maguire Kynch,
    Matthias Maier,
%     Jean-Paul Pelteret,
    Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

The class \texttt{VectorizedArray<Number>} is a key ingredient for the high
node-level performance of the matrix-free algorithms in deal.II~\cite{KronbichlerKormann2012, KronbichlerKormann2019}. It is a wrapper
class around a short vector of $n$ entries of type \texttt{Number} and maps
arithmetic operations to appropriate single-instruction/multiple-data (SIMD)
concepts by intrinsic functions.
The class \texttt{VectorizedArray} has been made more user-friendly in this release by making
it compatible with the STL algorithms found in the header \texttt{<algorithm>}.
The length of the  vector can now be queried by  \texttt{VectorizedArray::size()} and its underlying number type by \texttt{VectorizedArray::value\_type}.
Furthermore, the \texttt{VectorizedArray} class now supports range-based iterations over its entries.

Up to release 9.1, the
vector length $n$ has been set at compile time of the library to the highest
possible value supported by the given processor architecture.
Now, a second optional template argument
\texttt{VectorizedArray<Number, size>} can be given with \texttt{size} explicitly controlling
the vector length within the capabilities of a particular instruction set.
A full list of supported
vector lengths is presented in Table~\ref{tab:vectorizedarray}.
\todo[inline]{This table lacks the AltiVec support}

To account for the variable-size \texttt{VectorizedArray} class, all matrix-free related classes (like \texttt{MatrixFree} and \texttt{FEEvaluation})
have been extended with a new optional template argument specifying the
\texttt{VectorizedArrayType}. This allows users to select the vector length/ISA and,
as a consequence, the number of cells to be processed at once directly in their applications:
The deal.II-based
library \texttt{hyper.deal}~\cite{munch2020hyperdeal}, which solves the 6D Vlasov--Poisson equation with high-order
discontinuous Galerkin methods (with more than a thousand degrees of freedom per cell), constructs a tensor product of two \texttt{MatrixFree} objects of different SIMD-vector
length in the same application and benefits---in terms of performance---by the possibility of decreasing the number of cells processed by a single SIMD instruction.

@@

Furthermore, the new interfaces enable using any data structure
\texttt{VectorizedArrayType} as long as it supports required
functionalities like \texttt{size()} or \texttt{value\_type}. This prepares
for the \texttt{C++23} feature \texttt{std::simd} that will be enabled in the future.
Table~\ref{tab:simd} gives a comparison of the deal.II-specific SIMD classes and
the equivalent C++20 classes. Finally, this change also prepares for specialized
code paths exploiting
vectorization within an element~\cite{KronbichlerKormann2019} in the future.

%We welcome the standardization of the SIMD
%parallelization paradigm in C++ and intend to replace step by step our own
%wrapper class, which has been continuously developed over the last decade. We
%would like to emphasize that the work invested in this class was not in vain,
%since many performance-relevant utility functions implemented with \texttt{VectorizedArray} in mind (e.g., \texttt{vectorized\_load\_and\_transpose}
%and \texttt{vectorized\_transpose\_and\_store}) will be still used, since they
%have not become part of the standard.

@@

Large-scale simulations with 304,128 cores have revealed bottlenecks in setup
routines due to the usage of expensive collective operations
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak Alltoall()}, storing
problem sizes and offsets from all processors in the MPI universe.
In release 9.2, we have replaced these functions in favor of
consensus algorithms~\cite{hoefler2010scalable}, which can be
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}).
Now, only the locally relevant information about the index ranges is
(re)computed when needed, which, for more than 100 MPI processes, uses
point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

Users can apply the new algorithms for their own dynamic-sparse problems by
providing a list of target
processes and pack/unpack routines either by implementing the interface
\texttt{CA::\allowbreak Process} or by providing \texttt{std::function}
objects to \texttt{CA::AnonymousProcess}.
\todo[inline]{Aren't the target processes the result of the operation, and the
  input global numbers of indices that we request?}

By replacing the collective communications during setup and removing the arrays
that contain information for each process (enabled by the application of consensus
algorithms and other modifications---a full list of modifications leading to this
improvement can be found online), we were able to significantly improve the setup
time for large-scale simulations and to solve a Poisson problem with multigrid
with 12T unknowns.
\todo[inline]{Where did we solve for 12T unknowns? The largest I [Martin] have done is 2.1T unknowns.}
Figure~\ref{} compares the timings of a simulation (incl. setup) with the
previous release 9.1 and with the current release 9.2.
{\color{red}TODO[Peter/Martin] description of the results}
